A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-10-27 jour/jour.tex
more

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer \cite{blanchette-et-al-2013-smt}.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We
added two unpublished theories about Bird and Stern--Brocot trees
by Peter Gammie and Andreas Lochbihler (\gandl,
317~goals). To exercise the datatype support, theories about lists and trees
were added to the first two benchmark sets. The theories were selected before
conducting any experiments. The experimental data are available
online.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}

@@

For each goal in each theory, Sledgehammer selected about 16~lemmas,
which were monomorphized and translated to SMT-LIB~2 along with the goal.
The resulting problem was given to the
development version of CVC4 and to Z3 4.3.2 for comparison, each running for
up to 60~s on the StarExec cluster
\cite{stump-et-al-2014-starexec}.
Problems not involving any (co)datatypes were left out.
Due to the lack of machinery for reconstructing inferences about (co)datatypes
in Isabelle, the solvers are trusted as oracles. % in these experiments.
%Z3 is included for comparison.

The development version of CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured.
% to find
% out whether these somewhat expensive rules are useful in practice.
Even when the
decision procedure is disabled, the problems may contain
basic lemmas about constructors and selectors, allowing some (co)datatype
reasoning.
% when this makes sense.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}

@@

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 221 & 209 & 775 & 777 & 52 & 51 & 1048 & 1037 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 227 & -- & 780 & -- & 52 & -- & 1059 & -- \\
Full datatypes
  & 227 & 213 & 786 & 791 & 52 & 51 & 1065 & 1055 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 222 & -- & 804 & -- & 56 & -- & 1082 & -- \\
Full codatatypes
  & 223 & -- & 804 & -- & \bfseries 59 & -- & 1086 & -- \\%[\jot]
Full (co)datatypes
  & \bfseries 229 & -- & \bfseries 815 & -- & \bfseries 59 & -- & \bfseries 1103 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:benchsixteen}
\end{table*}

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 221 & 209 & 775 & 777 & 52 & 51 & 1048 & 1037 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 227 & -- & 780 & -- & 52 & -- & 1059 & -- \\
Full datatypes
  & 227 & 213 & 786 & 791 & 52 & 51 & 1065 & 1055 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 222 & -- & 804 & -- & 56 & -- & 1082 & -- \\
Full codatatypes
  & 223 & -- & 804 & -- & \bfseries 59 & -- & 1086 & -- \\%[\jot]
Full (co)datatypes
  & \bfseries 229 & -- & \bfseries 815 & -- & \bfseries 59 & -- & \bfseries 1103 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captionii}
\label{tab:benchtwofittysix}
\end{table*}

>>>>

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer \cite{blanchette-et-al-2013-smt}.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We
added unpublished theories about Bird and Stern--Brocot trees
by Peter Gammie and Andreas Lochbihler (\gandl,
317~goals). To exercise the datatype support, theories about lists and trees
were added to the first two benchmark sets. The theories were selected before
conducting any experiments. The experimental data are available
online.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}

@@

For each goal in each theory, we used Sledgehammer to select either 16 or
256~lemmas, which were monomorphized and translated to SMT-LIB~2 along with the
goal. The resulting problem was given to the
development version of CVC4 and to Z3 4.3.2 for comparison, each running for
up to 60~s on the StarExec cluster
\cite{stump-et-al-2014-starexec}.
Problems not involving any (co)datatypes were left out.
Due to the lack of machinery for reconstructing inferences about (co)datatypes
in Isabelle, the solvers are trusted as oracles. % in these experiments.
%Z3 is included for comparison.

The development version of CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured.
% to find
% out whether these somewhat expensive rules are useful in practice.
Even when the
decision procedure is disabled, the problems may contain
basic lemmas about constructors and selectors, allowing some (co)datatype
reasoning. This is especially the case for problems generated using 256~lemmas.
% when this makes sense.
%This allows us to answer the question, \relax{What are the
%benefits of activating the decision procedure as opposed to letting
%Sledgehammer do what it would normally do?}

@@

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 287 & 282 & 765 & 771 & 47 & 44 & 1099 & 1097 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 298 & -- & 771 & -- & 47 & -- & 1116 & -- \\
Full datatypes
  & 298 & 294 & 775 & 783 & 47 & 44 & 1120 & 1121 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 288 & -- & 797 & -- & 47 & -- & 1132 & -- \\
Full codatatypes
  & 288 & -- & 797 & -- & \bfseries 52 & -- & 1137 & -- \\%[\jot]
Full (co)datatypes
  & \bfseries 299 & -- & \bfseries 806 & -- & \bfseries 52 & -- & \bfseries 1157 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:benchsixteen}
\end{table*}

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 617 & 560 & 1503 & 1271 & 89 & 80 & 2209 & 1911 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 617 & -- & \bfseries 1504 & -- & 90 & -- & 2211 & -- \\
Full datatypes
  & 617 & 560 & \bfseries 1504 & 1263 & 90 & 78 & 2211 & 1901 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 617 & -- & 1501 & -- & 90 & -- & 2208 & -- \\
Full codatatypes
  & \bfseries 620 & -- & 1502 & -- & 98 & -- & \bfseries 2220 & -- \\%[\jot]
Full (co)datatypes
  & 619 & -- & 1501 & -- & \bfseries 99 & -- & 2219 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captionii}
\label{tab:benchtwofittysix}
\end{table*}
